---
title: "3: Stratified contrasts"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{3: Stratified contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(digits = 3)
```

```{r setup}
library(ratesci)
```

## Introduction to scoreci() for computing confidence intervals for stratified comparisons of independent binomial or Poisson rates

Stratified analysis might be required if a clinical trial employs stratified randomisation, or if results from multiple studies are to be combined in a meta-analysis. The illustration below uses data from a meta-analysis of 9 trials studying the effectiveness of graduated compression stockings for prevention of postoperative deep vein thrombosis (DVT) [from @roderick2005]:

```{r}
data(compress, package = "ratesci")
strat_rd <- scoreci(x1 = compress$event.gcs, 
                   n1 = compress$n.gcs, 
                   x2 = compress$event.control, 
                   n2 = compress$n.control, 
                   contrast = "RD", 
                   stratified = TRUE)
strat_rd$estimates
```

```{r}
strat_rd$pval
```

The Qtest object provides a heterogeneity test and related quantities, including an assessment of qualitative heterogeneity (indicating whether treatment effects are in opposite directions in different strata).

```{r}
strat_rd$Qtest
```

````{=html}
<!--
```{r}
strat_rr <- scoreci(x1 = compress$event.gcs, 
                   n1 = compress$n.gcs, 
                   x2 = compress$event.control, 
                   n2 = compress$n.control, 
                   contrast = "RR", 
                   stratified = TRUE)
strat_rr$Qtest
```
-->
````

The stratdata object provides estimates and intervals for each stratum, along with details of stratum weights and contributions to the Q-statistic (these may be further examined using the `hetplot` argument).

```{r}
strat_rd$stratdata
```

### Random effects

In this instance, there is weak evidence ($p = 0.07$) that the treatment effect (on the RD scale) varies across strata. (Estimation on the RR contrast scale gives more consistency across strata.) For the sake of illustration, a random effects analysis of RD would be obtained as follows, giving a wider interval that incorporates the stratum variability. The `random = TRUE` option invokes the t-distribution asymptotic score (TDAS) method [@laud2017], which modifies the asymptotic score methodology using formulae from the Hartung-Knapp-Sidik-Jonkman (HKSJ) random effects meta-analysis method (with superior performance compared to the DerSimonian-Laird method).

```{r}
strat_rd_rand <- scoreci(x1 = compress$event.gcs, 
                   n1 = compress$n.gcs, 
                   x2 = compress$event.control, 
                   n2 = compress$n.control, 
                   contrast = "RD", 
                   stratified = TRUE,
                   random = TRUE,
                   prediction = TRUE)
strat_rd_rand$estimates
```

Note that the TDAS method does not include a skewness correction, so when `random = TRUE`, the `skew` argument only affects the `stratdata` output element.

A prediction interval for the treatment effect in a new study [@higgins2008] can be obtained using `prediction = TRUE`:

```{r}
strat_rd_rand$prediction
```

## References
